---
title: "Lecture 8: Mediation Regression Analysis"
author: "Dr Nicolaas Puts and Dr Lucas França"
date: "28 November 2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background:

Job Search Intervention Study is a randomized study that investigates the efficacy of a job training intervention on unemployed workers. The program is designed to not only increase reemployment among the unemployed but also enhance their mental health. The data consists of a sample of n=899 subjects. Several variables have been recorded:

-	treat: indicator variable for whether participant was randomly selected for the job training programme (1= treatment group, 0= control group)
-	job_seek: a continuous scale measuring the level of job-search self-efficacy with values from 1 to 5.
-	depress: post treatment depression score
-	sex: gender (0 = male, 1 = female)
-	age: age (in years)



# Task 1

- Load the relevant packages for your analysis

```{r}
library(tidyverse)
library(foreign)
```

# Task 2

Load the dataset

```{r}
url <- 'https://github.com/fans-stats/intro_to_R_stats/blob/main/datasets/job_seek.sav?raw=true'

data <- read.spss(url, to.data.frame=TRUE) %>% 
  tibble::as_tibble()
data
```

# Task 3

- Create a new variable indicate whether the participant's age is above 40.

```{r}
data %>% 
  dplyr::mutate(above40 = dplyr::if_else(age > 40, TRUE, FALSE))
```

# Task 4:

- With summarise() estimate for the groups defined in task 4 the median age and depress and count how many received each type of treatment:

```{r}
data %>%  
  dplyr::mutate(above40 = dplyr::if_else(age > 40, TRUE, FALSE)) %>% 
  dplyr::group_by(above40) %>% 
  dplyr::summarise(med_age = median(age), 
                   med_depress = median(depress), 
                   total = length(above40),
                   treat_1_count = sum(treat)) %>% 
  dplyr::mutate(treat_0_count = total - treat_1_count)
```

# Task 5:

- Find the error in the following code section:

```{r}
dataset %>% 
  dplyr::mutate(sex = dplyr::recode(sex, 
                                    '0' = 'Male', 
                                    '1' = 'Female'))
  
  ggplot2::ggplot(aes(x = sex, y = depress)) + 
    geom_boxplot() + 
    geom_jitter() + 
    theme_bw()
```

Line 81: dataset -> data
Line 84: Missing %>% at the end

```{r}
data %>% 
  dplyr::mutate(sex = dplyr::recode(sex, 
                                    '0' = 'Male', 
                                    '1' = 'Female')) %>% 
  
  ggplot2::ggplot(aes(x = sex, y = depress)) + 
    geom_boxplot() + 
    geom_jitter() + 
    theme_bw()
```