---
title: "Lecture 2 – Importing and handling data I"
author: "Dr Nicolaas Puts and Dr Lucas França"
date: "10 October 2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Data types

The R data types or classes are

 - logical (or boolean)
 - integer
 - numeric
 - complex
 - character
 - raw (we are not going to use this one in the course)

## 1. Logical

```{r}
logic1 <- TRUE
logic2 <- FALSE
```


## 2. Integer

```{r}
int1 <- 16
int2 <- 8
```


## 3. Numeric

```{r}
num1 <- 3.5
num <- 6.1
```


## 4. Complex

```{r}
comp1 <- 2 + 5i
```


## 5. Character

```{r}
fruit <- "Apple and Banana"
```


## 6. Raw

```{r}
raw_ <- charToRaw("Apple")
```



# Data structure

## String

```{r}
fruit1 <- "Apple"
fruit2 <- "Banana"
fruit3 <- "Grapes"

nchar(fruit1)

paste(fruit1, fruit2, fruit3)
```
```{r}
fruit1 == fruit1
```


## Vectors

```{r}

fruits_vec <- c("Apple", "Banana", "Grapes")
fruits_vec[2]
```
```{r}
fruits_vec2 <- c(fruit1, fruit2, fruit3)
fruits_vec2
```

```{r}
num_vec <- c(1,5,6,7,2,3,5)
```

```{r}
seq_vec <- c(1:12)
seq_vec
```


## Matrix

```{r}
matrix1 <- matrix(c(1:6), nrow = 2, ncol = 3, byrow = TRUE)
matrix1
```

```{r}
matrix1[1,2]
```

```{r}
matrix1[1,]
```
```{r}
matrix1[,3]
```


## Arrays

```{r}
array1 <- array(c(1:12), dim = c(2,3,2))
array1
```


## Dataframe

```{r}
dataframe1 <- data.frame(
  Name = c("Anne", "Matthew", "Simon"),
  Age = c(21, 18, 19),
  Vote = c(TRUE, FALSE, TRUE)
)

dataframe1
```

```{r}
table(dataframe1$Vote)
```


## Tibbles

```{r}
library(tidyverse)
```

```{r}
tibble1 <- tibble(
  Name = c("Anne", "Matthew", "Simon"),
  Age = c(21, 18, 19),
  Vote = c(TRUE, FALSE, TRUE)
)
```


# Loading data

## Loading csv

```{r}
urlfile <- 'https://raw.githubusercontent.com/HeJasonL/BIG-TAC/main/TutorialData.csv' 
```

```{r}
dataset <- read.csv(urlfile)
dataset
```

```{r}
table(dataset$Sex)
```


## Loading SPSS

```{r}
library(foreign)
```


```{r}
urlfile <- 'https://github.com/fans-stats/intro_to_R_stats/blob/main/datasets/researchers_sample.SAV?raw=true'
```


```{r}
dataset2 <- read.spss(urlfile)
dataset2 <- as.tibble(dataset2)
dataset2
```


# A simple task

There's a sample of 25 people where the mean height is 180 cm with an SD of 30. Use R to calculate the 95% CI and put the CI into a variable.

```{r}
LL <- 180 - 1.96 * (30/sqrt(25))
UL <- 180 + 1.96 * (30/sqrt(25))

CI <- c(LL, UL)
CI
```

# Handling missing data

```{r}
dataset$Sensory1[37] <- NA
```

```{r}
table(is.na(dataset$Sensory1))
```

```{r}
min(dataset$Sensory1, na.rm = TRUE)
```
```{r}
mean(dataset$Sensory1, na.rm = TRUE)
```

```{r}
sd(dataset$Sensory1, na.rm = TRUE)
```

```{r}
which(is.na(dataset$Sensory1))
```

